{\includegraphics[width=3cm, height=2.5cm]{Cran.jpg}} \\ \includegraphics[width=7cm, height=1cm]{TB.jpg} \\ {Technical Documentation for TersectBrowser+}

### 2025-05-05 TechDoc/TechDoc.tex

The extensions were designed and added by the named authors under the supervision of of Dr Tomasz Kurowski between March and May 2025.

@@

\subsection{Embedding The Extensions }

@@

The genomebrowser relies on \texttt{Jbrowse} [TODO: Reference] extensively. Since Jbrowse is written in \texttt{React} [TODO: Reference], it is embedded in the new angular app via the \texttt{ReactComponentDirective} provided by the \texttt{@ngeenx/ngx-react} package on NPM.

@@

\begin{itemize}
    \item (1) open a genome browser within Tersect Browser and investigate specific variants;
    \item (2) see specific genes of interests against suspected introgressions; 
    \item (3) identify areas where high-impact variants are visible; 
    \item (4) be able to genereate DNA barcodes for the varioius cultivars and species in the dataset; and,  
    \item (5) to be able to load additional datasets, including for other crops. 
\end{itemize}

@@

The main requirements for TersectBrowser+ deployment are:
\begin{itemize}
    \item nvm (versions 16 and 22 specifically, as these will be deployed separately)
    \item npm registery \url{https://registry.npmjs.org/}
    \item Angular CLI v1.7.1
    \item Tersect CLI
    \item JBrowse CLI
    \item MongoDB
    \item Python virtual environment
    \item RapidNJ / Rosetta
\end{itemize}

@@

        \verb+python setup_new_tbrowser_dataset.py -f ${fasta} -g ${gff} -V ${vcfs}+
        \begin{itemize}
            \item After running the script, there should be visible a new \verb+config.json+ file in the root folder, as well as a new folder \verb+./~/mongo-data/gp_data_copy+ containing fasta files, gff files, and VCF files along with their index files. The main Tersect index file should be in the folder above. This is where the server will look to find information on tracks when generating the Variant Browser panel.
            \item This script checks for the install of modules such as SAMtools, BCFtools, and HTSlib, and will attempt to download these if not found. If a multi-sample VCF file is provided, it will split this file into individual sample level VCFs before proceeding. It will make sure these files are unzipped before using TERSECT CLI to build a tersect index for the whole dataset (it sources a previously created .tersect virtual environment to run this). After recompressing and indexing the fasta and VCF files, it will edit the \verb+add_example_dataset.sh+ script to point to the correct index file, and begin dataset addition to TersectBrowser+. Lastly, it will add the GFF file as a gene models track accessible within the browser.
        \end{itemize}
\end{enumerate}

\subsection{Deployment with loaded dataset}
\begin{enumerate}
    \item Run the Mongo Database using: \\
        \verb+mongod --dbpath <pathto/mongo-data>+
    \item Change to the root of the Tersect-browser repo and run:\\
        \verb+nvm use 16+\\
        \verb+npm start+
    \item Change to the ersect-browser/extension/genome-browser/ directory and run:\\
        \verb+nvm use 22+\\
        \verb+npm start+
    \item Open TersectBrowser+ on \url{http://localhost:4200/TersectBrowserGP/}
\end{enumerate}

@@

\textit{Tersect Browser+} comprises Tersect CLI, Tersect Browser and four extensions that meet the \textit{Teresect Browser+} use case in Section 3. These were:

@@

On reviewing node-compatible browser tools such as CRAMER, JBrowse was selected as the prefered solution, due to its high customisability, up-to-date features and active technical support. The initial build of JBrowse into \textit{TersectBrowser+}, focused on retaining features most relevant to the user aims. This includeed bringing the styling of the panel in line with the rest of the browser page. The extension also involved generating a popup window of the same JBrowse component to be initialised when the user identified a specific accession or bin region of interest.

@@

\subsubsection{Main Window Location and Tracks}
The main window is located at the top of Tersect Browser, below the settings bar but above the heatmap. Main window shows only sequence tracks and linear genome scale. Interactive features such as zoom anc chromosome selection have been removed from the main window. The container window has a dynamic offset, which changes according to tree view and zoom level to ensure the chromosome scales between TersectBrowser and JBrowse line up. By default, the main window shows only the gene models track.

The tracks include: 
\begin{itemize}
    \item Reference sequence 
    \item Variant tracks for each loaded accession
    \item Gene models FeatureTrack (\nameref{sec:Models})
\end{itemize}

@@

\subsubsection{Jbrowser TersectBrowser Syncing}

@@

\paragraph{View State} 
The \verb +JBrowseLinearGenomeView+ component is imported from @jbrowse/react-linear-genome-view and is wrapped using React. The configurations for the assembly, assembly, and variant tracks, and tracks, are imported from the respective assembly.ts and tracks.ts configuration files. The styling configurations, config, are imported from jbrowseConfig.ts. The \verb +JbrowseWrapperProps+ are imported from \verb +JbrowseInterface+.

@@

\subsection{Feature Search Extension - Technical Description}

@@

@@

\subsubsection{Handling Search Request}

An API endpoint was added to the \texttt{tersect-router} backend to support search functionality. The structure of the endpoint is as follows:

@@

\subsubsection{Handling Searching by Interval}

@@

To associate variants with accession names, a function uses \texttt{tersect} to determine which accessions have each specific variant. A function to get the optimal concurrency (\texttt(getOptimalConcurrency)) was introduced which detected the number of cores present on the server, subtracting 1 from the total and using the rest. To prevent server overload from concurrent \texttt{tersect} calls, the \texttt{p-limit} library is used to control concurrency based on available CPU cores.

@@

\subsubsection{Handling Searching by Gene Name}

@@

Variant searches can also be triggered from:
\begin{itemize}
  \item Each bin on the bin plot.
  \item A selected range on the scale plot.
\end{itemize}

These triggers emit a custom event \texttt{tersect-search-variants}, which sets the interval and opens the modal for user interaction.

@@

\subsubsection{VCF Mechanism Summary}

The overall mechanism for variant search begins with API queries, uses either \texttt{bcftools} or Trix for data retrieval, links variants to accessions with \texttt{tersect}, and finally triggers the highlighting of bins in the Tersect UI.

\paragraph{Other Triggers for Variant Search}
Variant search also triggered via bin plot or scale plot selection using the \texttt{tersect-search-variants} custom event.

>>>>

The extensions were designed and added by the named authors under the supervision of Dr Tomasz Kurowski between March and May 2025.

@@

\subsection{Embedding The Extensions}

@@

The genome browser relies on \texttt{Jbrowse} [TODO: Reference] extensively. Since Jbrowse is written in \texttt{React} [TODO: Reference], it is embedded in the new angular app via the \texttt{ReactComponentDirective} provided by the \texttt{@ngeenx/ngx-react} package on NPM.

@@

\begin{itemize}
    \item (1) open a genome browser within Tersect Browser and investigate specific variants;
    \item (2) see specific genes of interests against suspected introgressions; 
    \item (3) identify areas where high-impact variants are visible; 
    \item (4) be able to generate DNA barcodes for the various cultivars and species in the dataset; and,  
    \item (5) to be able to load additional datasets, including for other crops. 
\end{itemize}

@@

The main requirements for TersectBrowser+ deployment are:
\begin{itemize}
    \item nvm (versions 16 and 22 specifically, as these will be deployed separately)
    \item npm registry \url{https://registry.npmjs.org/}
    \item Angular CLI v1.7.1
    \item Tersect CLI
    \item JBrowse CLI
    \item MongoDB
    \item Python virtual environment
    \item RapidNJ / Rosetta
\end{itemize}

@@

        \verb+python setup_new_tbrowser_dataset.py -f ${fasta} -g ${gff} -V ${vcfs}+
        \begin{itemize}
            \item After running the script, there should be visible a new \verb+config.json+ file in the root folder, as well as a new folder \verb+./~/mongo-data/gp_data_copy+ containing fasta files, gff files, and VCF files along with their index files. The main Tersect index file should be in the folder above. This is where the server will look to find information on tracks when generating the Variant Browser panel.
            \item This script checks for the installation of modules such as SAMtools, BCFtools, and HTSlib, and will attempt to download these if not found. If a multi-sample VCF file is provided, it will split this file into individual sample level VCFs before proceeding. It will make sure these files are unzipped before using TERSECT CLI to build a tersect index for the whole dataset (it sources a previously created tersect virtual environment to run this). After recompressing and indexing the fasta and VCF files, it will edit the \verb+add_example_dataset.sh+ script to point to the correct index file, and begin dataset addition to TersectBrowser+. Lastly, it will add the GFF file as a gene models track accessible within the browser.
        \end{itemize}
\end{enumerate}

\subsection{Deployment with loaded dataset}
\begin{enumerate}
    \item Run the Mongo Database using: \\
        \verb+mongod --dbpath <pathto/mongo-data>+
    \item Change to the root of the Tersect-browser repo and run:\\
        \verb+nvm use 16+\\
        \verb+npm start+
    \item Change to the Tersect-browser/extension/genome-browser/ directory and run:\\
        \verb+nvm use 22+\\
        \verb+npm start+
    \item Open TersectBrowser+ on \url{http://localhost:4200/TersectBrowserGP/}
\end{enumerate}

@@

\textit{Tersect Browser+} comprises Tersect CLI, Tersect Browser and four extensions that meet the \textit{Tersect Browser+} use case in Section 3. These were:

@@

On reviewing node-compatible browser tools such as CRAMER, JBrowse was selected as the preferred solution, due to its high customisability, up-to-date features and active technical support. The initial build of JBrowse into \textit{TersectBrowser+}, focused on retaining features most relevant to the user aims. This included bringing the styling of the panel in line with the rest of the browser page. The extension also involved generating a popup window of the same JBrowse component to be initialised when the user identified a specific accession or bin region of interest.

@@

\subsubsection{Main Window Location and Tracks}
The main window is located at the top of Tersect Browser, below the settings bar but above the heatmap. Interactive features such as zoom and chromosome selection have been removed from the main window. The container window has a dynamic offset, which changes according to tree view and zoom level to ensure the chromosome scales between TersectBrowser and JBrowse line up. By default, the main window shows only the gene models track.

@@

\subsubsection{Jbrowse TersectBrowser Syncing}

@@

\paragraph{View State} 
The \verb +JBrowseLinearGenomeView+ component is imported from \verb*|@jbrowse/react-linear-genome-view| and is wrapped using React. The configurations for the assembly, assembly, and variant tracks, and tracks, are imported from the respective assembly.ts and tracks.ts configuration files. The styling configurations, config, are imported from jbrowseConfig.ts. The \verb +JbrowseWrapperProps+ are imported from \verb +JbrowseInterface+.

@@

@@

The preparation process is automated with the \verb*|{build_annotate_per_chrom.sh}| script (Supplementary File 1). Its example usage is;

\begin{verbatim}
    build_annotate_per_chrom.sh tracks.json index.tsi genes.gff3.gz [outdir]
\end{verbatim}

Where \texttt{tracks.json} is the JBrowse tracks configuration file (TODO: Reference), \texttt{index.tsi} is the Tersect index file, \texttt{genes.gff3.gz} is the compressed GFF3 gene annotation file, and \texttt{[outdir]} is the output directory for the generated per-chromosome VCF and annotation files.

@@

The Trix file generation is automated with the \verb*|(tersect_trix_builder.sh)| script (Supplementary File 2).

\subsubsection{Handling A Search Request}

An API endpoint was added to the \texttt{tersect-router} file on the backend to support search functionality. The structure of the endpoint is as follows:

@@

\subsubsection{Searching by Interval}

@@

To associate variants with accession names, tersectForKey function uses \texttt{tersect} to determine which accessions have each specific variant. A function to get the optimal concurrency \texttt(getOptimalConcurrency) was introduced which detected the number of cores present on the server, subtracting 1 from the total and using the rest. To prevent server overload from concurrent \texttt{tersect} calls, the \texttt{p-limit} library is used to control concurrency based on available CPU cores.

@@

\subsubsection{Searching by Gene Name}

@@

Variant search can also be triggered from:
\begin{itemize}
  \item Each bin on the bin plot.
  \item A selected range on the scale plot.
\end{itemize}

These triggers emit a custom event \texttt{tersect-search-variants}, which is being listened for on the react component. This sets the interval in the search dialog and opens the modal for user interaction.

@@
